---
title: "Class 11: AlphaFold"
author: "Areidy Arroyo A17412951"
format: gfm
toc: TRUE
---

## Background

In this hands-on session we will utilize AlphaFold to predict protein structure from sequence.

Without the aid of such approaches, it can take years of expensive laboratory work to determine the structure of just one protein. With AlphaFold we can now accurately compute a typical protein structure in as little as ten minutes.

The PDB database (the main repository of experimental structures) only has **~250 thousand** structures (we saw this in the last lab). The main protein sequence database over **200 million** sequences! Only 0.125% of known sequences have known structure - this is called the "structure knowledge gap".

```{r}
(250000 / 200000000 ) *100
```

- Structures are much harder to determine than sequences
- They are expensive (on average ~$1 million each)
- They take on average 3-5 years to solve!

## EBI AlphaFold Database

The EBI has a database of pre-computed ALphaFold (AF) models called AFDB.
This is growing overtime and can be useful to check before running AF ourselves

## Running AlphaFold

We can download and run locally (on our own computers) but we need a GPU. Or we can use "cloud" computing to run this on someone elses computers :-)

We will use ColabFold < https://github.com/sokrypton/ColabFold >

We previously found there was no AFDB entry for our HIV sequence:

```
>HIV-Pr-Dimer
PQITLWQRPLVTIKIGGQLKEALLDTGADDTVLEEMSLPGRWKPKMIGGIGGFIKVRQYDQILIEICGHKAIGTVLVGPTPVNIIGRNLLTQIGCTLNF:PQITLWQRPLVTIKIGGQLKEALLDTGADDTVLEEMSLPGRWKPKMIGGIGGFIKVRQYDQILIEICGHKAIGTVLVGPTPVNIIGRNLLTQIGCTLNF
```

Here we will use AlphaFold2_mmseqs2

## Custom analysis of resulting models

We will move our AlphaFold results into RStudio

```{r}
results_dir <- "hivpr_23119/" 
```

```{r}
pdb_files <- list.files(path=results_dir,
                        pattern="*.pdb",
                        full.names = TRUE)
basename(pdb_files)
```

Here we will see the model sequences and all its alignments

```{r}
library(bio3d)

# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```
We will now calculate the structural distance betwee coordinate sets using RMSD

```{r}
rd <- rmsd(pdbs, fit=T)

range(rd)
```

Create a pheatmap of these coordinates

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```
We can now plot the pLDDT values into a PDB files and add our desired colors into the graph

```{r}
pdb <- read.pdb("1hsg")
```

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

Fix the superposition of the model creating a rigid core using `core.find()`

```{r}
core <-core.find(pdbs)
core.inds <- print(core, vol=0.5)
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")

```

Now that we have superposed coordinates we can open our structure from Mol*

Here we will examine RMSF and its positions 

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```
## Predicted Alignment Error for domains

Read files and predict alignment error for domains using `jsonlite()`

```{r}
library(jsonlite)

pae_files <- list.files(path=results_dir,
                       pattern=".*model.*\\.json",
                        full.names = TRUE)
```

For example we will read models 1 and 5 here:

```{r}
pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae1)
```

```{r}
head(pae1$plddt) 
```

Calculate the max PAE value 

```{r}
pae1$max_pae
pae5$max_pae
```

We can now plot this

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

Now we will plot using the same z range
```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```


## Residue conservation from alignment file

```{r}
aln_file <- list.files(path=results_dir,
                       pattern=".a3m$",
                        full.names = TRUE)
aln_file
```

```{r}
aln <- read.fasta(aln_file[1], to.upper = TRUE)
```

Calculate how many sequences are in the alignment

```{r}
dim(aln$ali)
```

Use `conserv()` to score the residue conservation

```{r}
sim <- conserv(aln)
plotb3(sim[1:99], sse=trim.pdb(pdb, chain="A"),
       ylab="Conservation Score")
```

View parts where there is a high cutoff value in the consensus sequence

```{r}
con <- consensus(aln, cutoff = 0.9)
con$seq
```

```{r}
m1.pdb <- read.pdb(pdb_files[1])
occ <- vec2resno(c(sim[1:99], sim[1:99]), m1.pdb$atom$resno)
write.pdb(m1.pdb, o=occ, file="m1_conserv.pdb")
```


