---
title: "Class12"
author: "Areidy Arroyo (A17412951)"
format: gfm
toc: TRUE
---

## Section 1. Proportion of G/G in a population

Download CSV file from Ensemble containing MXL from sample rs8067378

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

To look for Great Britain we will look at the GBR population,still using the CSV file we provided from earlier

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

We will now find the proportion of G | G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

## Section 4: Population Scale Analysis

We will be using multiple samples in order to be able to observe the popultaion and its genetic differences based on population scale.

We have processed about 230 saomes and wille examine this to find its relationship between rs8067378 and/or ORMDL3 expression

> Q13. Read this file into R and determine the sample size for each genotype and
their corresponding median expression levels for each of these genotypes

Total samples

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
tapply(expr$exp, expr$geno, median)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the
relative expression value between A/A and G/G displayed in this plot? Does the
SNP effect the expression of ORMDL3?

Generate a box plot using `ggplot2()`

```{r}
library(ggplot2)

ggplot(expr) + aes(x=geno, exp, fill=geno) +
geom_boxplot(notch=TRUE)
```

We can see that the SNP does affect ORMDL3 and its expression. We can see that the genotypes A/A A/G and G/G all appear at different expressions, some higher than others. Based on the boxplot we have created the A/A genotype is much highly expressed than the others. We can notice that G tends to decrease expression meanwhile A will increase expression.


