---
title: "Class 17: Analyzing Sequence Data in Cloud"
author: "Areidy Arroyo (A17412951)"
format: gfm
toc: TRUE
---

## Downstream Analysis

Install R package `BiocManager::install("tximport")` used to import transcript level RNA sequence results. We will be using gene counts during downstream analysis

```{r}
library(tximport)

# setup the folder and file-names to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

Provide transcript counts for each sample

```{r}
head(txi.kallisto$counts)
```

Total count calculated for each sample

```{r}
colSums(txi.kallisto$counts)
```

Filter transcripts with no reads

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```


## Principal Component Analysis

Summarize PCA samples 

```{r}
pca <- prcomp(t(x), scale=TRUE)
```

```{r}
summary(pca)
```

PC1 vs PC2

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

PC1 vs PC3

```{r}
plot(pca$x[,1], pca$x[,3],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC3", pch=16)
```

PC2 vs PC3

```{r}
plot(pca$x[,2], pca$x[,3],
     col=c("blue","blue","red","red"),
     xlab="PC2", ylab="PC3", pch=16)
```


