---
title: "Class06: R Functions"
author: "Areidy (PID: A17412951)"
format: pdf
toc: true
---

## Background

Functions are at the heart of using R. Everything we do involves calling and using function (from data input, analysis to results output)

All functions in R have at least 3 things:

1. A **name** the thing we use to call the function.
2. One or more input **arguments** that a are comma separated
3. The **body**, lines of code between curly brackets { } that does the work of the function.

## A first function

Let's write a silly wee function to add some number:

```{r}
add <- function(x) {
  x + 1
}
```

Let's try it out
```{r}
add(100)
```

Will this work
```{r}
add(c(100, 200, 300))
```

Modify to be more useful and add more than just 1

```{r}
add <- function(x, y=1) {
  x + y
}
```

```{r}
add(100, 10)
```

Will this still work
```{r}
add(100)
```

```{r}
plot(1:10, col="blue", typ="b")
```

```{r}
log(10, base=10)
```

> **N.B.** Input arguments can be either **required** or **optional**. The later have a fall-back default that is specified in the function code with an equals sign.


```{r}
#add(x=100, y=200, z=300)
```


## A second function

All functions in R look like this

```
name <- function(arg) { 
  body 
}
```

The `sample()` function in R ...

```{r}
sample(1:10, size = 4)
```

> Q. Return 12 numbers picked randomly from the input 1:10

```{r}
sample(1:10, size = 12, replace = TRUE)
```

> Q. Write the code to generate a random 12 nucelotide long DNA sequence?

```{r}
bases <- c("A", "C", "G", "T")
sample(bases, size=12, replace=TRUE)
```


> Q. Write a first version function called `generate_dna()` that generates a user specified length `n` random DNA sequence? 

```{r}
generate_dna <- function(n=6) { 
  bases <- c("A", "C", "G", "T")
  sample(bases, size = n, replace = TRUE)
}
```

```{r}
generate_dna(100)
```

> Q. Modify your funtion to return a FASTA like sequence so rather than [1] "T" "C" "T" "G" "C" we want "TCTGC"

```{r}
generate_dna <- function(n=6) { 
  bases <- c("A", "C", "G", "T")
  ans <- sample(bases, size = n, replace = TRUE)
  ans <- paste(ans, collapse = "")
  return(ans)
  x <- "pooopoppants"
}
```

```{r}
generate_dna(10)
```


An example

```{r}
# Example pattern (not using your bases)
x <- c("T", "C", "T", "G", "C")

paste(x, collapse = "TCTGC")
# returns "HELLO"
```

>. Q. Give the user an option to return to FASTA format output sequence or standard multi-elemnt vector format?

```{r}
generate_dna <- function(n=6, fasta= TRUE) { 
  bases <- c("A", "C", "G", "T")
  ans <- sample(bases, size = n, replace = TRUE)
  
  if (fasta) { 
    ans <- paste(ans, collapse = "") 
    cat("Hello...")
  } else {
      cat("...is it me you are looking for...")
  }

  return(ans)
}
```

```{r}
generate_dna(10)
generate_dna(10, fasta = F)
```

> Q. Write a function called `generate_protein()` that generates a user specified length protein sequence in FASTA like format?

```{r}
generate_protein <- function(n){
  
aa <- c("A", "R", "N", "D", "C", "Q", "E", "G", "H", "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")
  
  ans <- sample(aa, size = n, replace = T)
  ans <- paste (ans, collapse= "")
  return(ans)
}
```


```{r}
generate_protein(10)
```



> Q. Use your new `generate_protein()` function to generate all sequences length 6 and 12 amin-acids in length and check of any of these are unique in nature(i.e. found in the NR database at NCBI)?

```{r}
generate_protein(6)
generate_protein(7)
generate_protein(8)
generate_protein(9)
generate_protein(10)
generate_protein(11)
generate_protein(12)

```

or we could do a `for()` loop:

```{r}
for(i in 6:12) {
  cat(">", i, sep= "", "\n")
  cat( generate_protein(i), "\n")
}
```

>id
AGKRTST
>next
G


N.B> 6-8 sequences were found on NCBI